---
title: "p8105_hw2_mts2203"
author: "Madeline Sharp"
date: "2026-10-03"
output: github_document
---

# Problem 1

## Loading libraries

``` {r load_libraries, message = F}  
# loading tidyverse library
library(tidyverse)
```

## Reading and cleaning the Problem 1 data

``` {r reading_and_cleaning_p1_data, message = F}  
# importing and cleaning the NYC Transit csv data
nyc_transit_df = 
  read_csv(
    file = paste0(
      "p8105_hw2_mts2203_files/",
      "NYC_Transit_Subway_Entrance_And_Exit_Data.csv"), 
    na = c("", "NA", ".")
    ) |>
  janitor::clean_names() |>
  select(line, station_name, station_latitude,  station_longitude, 
         starts_with("route"), 
         entry, vending, entrance_type, ada
         ) |>
  mutate (
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )
```

## Short paragraph about the dataset

The `NYC_Transit_Subway_Entrance_And_Exit_Data.csv` dataset contains variables `r paste(names(read_csv("p8105_hw2_mts2203_files/NYC_Transit_Subway_Entrance_And_Exit_Data.csv")), collapse = ", ")`. I created data frame `nyc_transit_df` to import and clean the data from the `NYC_Transit_Subway_Entrance_And_Exit_Data.csv` dataset. First, I used `read_csv()` read the csv file, second, I standardized the column names using `janitor::clean_names()`, third, I selected for the columns `r paste(names(nyc_transit_df), collapse = ", ")`, and finally I used the `mutate()` function to change the `entry` column data type from `character` to `logical`. The function `case_match()` was essential to accomplishing the last task. The resulting `nyc_transit_df` dataset's dimensions are`r nrow(nyc_transit_df)` rows by `r ncol(nyc_transit_df)` columns. No, this data is not yet tidy. The route columns should be better consolidated, the vending column could be a logical instead of a character data type, and the missing values have not been dealt with.

## Addressing questions

``` {r answering_questions, message = F}  
# finding the number of distinct stations
nyc_transit_df |>
  distinct(station_name, line) |> 
  nrow()

# finding the number of stations that are ADA compliant
nyc_transit_df |>
  distinct(station_name, line, ada) |>
  filter(ada == TRUE) |>
  nrow()

# finding the proportion of entrances/exits w/o vending that allow entrance
nyc_transit_df |>
  filter(vending == "NO") |>
  summarise(proportion = mean(entry, na.rm = TRUE))
```

## Reformatting data

``` {r reformatting_data, message = F}  
# reformatting data to make route number ad route name distinct variables
nyc_transit_routes = 
  nyc_transit_df |>
  mutate(
    across(
      starts_with("route"), as.character
      )
  ) |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
  ) |>
  drop_na(route_name)
```

## Answering questions with reformatted data

``` {r answering_route_questions_with_reformatted_data, message = F}  
# finding the number of distinct stations that serve the A train
nyc_transit_routes |>
  distinct(station_name, line, route_name) |>
  filter(route_name == "A") |>
  nrow()

# finding the number of ADA compliant stations that serve the A train
nyc_transit_routes |>
  distinct(station_name, line, route_name, ada) |>
  filter(route_name == "A", ada == TRUE) |>
  nrow()
```


# Problem 2

## Reading and cleaning the Problem 2 data

```{r reading_and_cleaning_p2_data, message = F}
#importing and cleaning the Mr. Trash Wheel excel sheet data
mr_tw_df =
  readxl::read_excel(
    paste0(
      "p8105_hw2_mts2203_files/",
      "202509 Trash Wheel Collection Data.xlsx"),
    sheet = "Mr. Trash Wheel",
    range = "A2:N710",
    na = c("", "NA", "."),
    skip = 1
  ) |>
  janitor::clean_names() |>
  filter(
    !is.na(dumpster)
  ) |>
  mutate(
    sports_balls = as.integer(round(sports_balls))
  )
```

## Reading, cleaning and organizing more Problem 2 data

```{r organizing_other_p2_data, message = F}
#importing and organizing the Professor Trash Wheel excel sheet data
prof_tw_df = 
  readxl::read_excel(
    paste0(
      "p8105_hw2_mts2203_files/",
      "202509 Trash Wheel Collection Data.xlsx"),
    sheet = "Professor Trash Wheel",
    range = "A2:M135",
    na = c("", "NA", "."),
    skip = 1
  ) |>
  janitor::clean_names() |>
  filter(
    !is.na(dumpster)
  )

#importing and organizing the Gwynn Falls Trash Wheel excel sheet data
gwynns_falls_tw_df = 
  readxl::read_excel(
    paste0(
      "p8105_hw2_mts2203_files/",
      "202509 Trash Wheel Collection Data.xlsx"),
    sheet = "Gwynns Falls Trash Wheel",
    range = "A2:L352",
    na = c("", "NA", "."),
    skip = 1
  ) |>
  janitor::clean_names() |>
  filter(
    !is.na(dumpster)
  )
```