---
title: "p8105_hw2_mts2203"
author: "Madeline Sharp"
date: "2026-10-03"
output: github_document
---

# Problem 1

## Loading libraries

``` {r load_libraries, message = F}  
# loading tidyverse library
library(tidyverse)
```

## Reading and cleaning the data

``` {r reading_data, message = F}  
# importing anc cleaning the NYC Transit csv data
nyc_transit_df = 
  read_csv(
    file = paste0(
      "p8105_hw2_mts2203_files/",
      "NYC_Transit_Subway_Entrance_And_Exit_Data.csv"), 
    na = c("", "NA", ".")
    ) |>
  janitor::clean_names() |>
  select(line, station_name, station_latitude,  station_longitude, route1, 
         route2, route3, route4, route5, route6, route7, route8, route9, 
         route10, route11, entry, vending, entrance_type, ada) |>
  mutate (
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )
```

## Short paragraph about the dataset

The `NYC_Transit_Subway_Entrance_And_Exit_Data.csv` dataset contains variables `r paste(names(read_csv("p8105_hw2_mts2203_files/NYC_Transit_Subway_Entrance_And_Exit_Data.csv")), collapse = ", ")`. I created data frame `nyc_transit_df` to import and clean the data from the `NYC_Transit_Subway_Entrance_And_Exit_Data.csv` dataset. First, I used `read_csv()` read the csv file, second, I standardized the column names using `janitor::clean_names()`, third, I selected for the columns `r paste(names(nyc_transit_df), collapse = ", ")`, and finally I used the `mutate()` function to change the `entry` column data type from `character` to `logical`. The function `case_match()` was essential to accomplishing the last task. The resulting `nyc_transit_df` dataset's dimensions are`r nrow(nyc_transit_df)` rows by `r ncol(nyc_transit_df)` columns.