---
title: "p8105_hw2_mts2203"
author: "Madeline Sharp"
date: "2026-10-03"
output: github_document
---

# Problem 1

## Loading libraries

``` {r load_libraries, message = F}  
# loading tidyverse library
library(tidyverse)
```

## Reading and cleaning the Problem 1 data

``` {r reading_and_cleaning_p1_data, message = F}  
# importing and cleaning the NYC Transit CSV data
nyc_transit_df = 
  read_csv(
    file = paste0(
      "p8105_hw2_mts2203_files/",
      "NYC_Transit_Subway_Entrance_And_Exit_Data.csv"), 
    na = c("", "NA", ".")
    ) |>
  janitor::clean_names() |>
  select(line, station_name, station_latitude,  station_longitude, 
         starts_with("route"), 
         entry, vending, entrance_type, ada
         ) |>
  mutate (
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )
```

## Short paragraph about the dataset

The `NYC_Transit_Subway_Entrance_And_Exit_Data.csv` dataset contains variables `r paste(names(read_csv("p8105_hw2_mts2203_files/NYC_Transit_Subway_Entrance_And_Exit_Data.csv")), collapse = ", ")`. I created data frame `nyc_transit_df` to import and clean the data from the `NYC_Transit_Subway_Entrance_And_Exit_Data.csv` dataset. First, I used `read_csv()` to read the CSV file, second, I standardized the column names using `janitor::clean_names()`, third, I selected for the columns `r paste(names(nyc_transit_df), collapse = ", ")`, and finally I used the `mutate()` function to change the `entry` column data type from `character` to `logical`. The function `case_match()` was essential to accomplishing the last task. The resulting `nyc_transit_df` dataset's dimensions are`r nrow(nyc_transit_df)` rows by `r ncol(nyc_transit_df)` columns. No, this data is not yet tidy. The route columns should be better consolidated, the vending column could be a logical instead of a character data type, and the missing values have not been dealt with.

## Addressing questions

``` {r answering_questions, message = F}  
# finding the number of distinct stations
nyc_transit_df |>
  distinct(station_name, line) |> 
  nrow()

# finding the number of stations that are ADA compliant
nyc_transit_df |>
  distinct(station_name, line, ada) |>
  filter(ada == TRUE) |>
  nrow()

# finding the proportion of entrances/exits w/o vending that allow entrance
nyc_transit_df |>
  filter(vending == "NO") |>
  summarise(proportion = mean(entry, na.rm = TRUE))
```

## Reformatting data

``` {r reformatting_data, message = F}  
# reformatting data to make route number and route name distinct variables
nyc_transit_routes = 
  nyc_transit_df |>
  mutate(
    across(
      starts_with("route"), as.character
      )
  ) |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
  ) |>
  drop_na(route_name)
```

## Answering questions with reformatted data

``` {r answering_route_questions_with_reformatted_data, message = F}  
# finding the number of distinct stations that serve the A train
nyc_transit_routes |>
  distinct(station_name, line, route_name) |>
  filter(route_name == "A") |>
  nrow()

# finding the number of ADA compliant stations that serve the A train
nyc_transit_routes |>
  distinct(station_name, line, route_name, ada) |>
  filter(route_name == "A", ada == TRUE) |>
  nrow()
```


# Problem 2

## Reading and cleaning the Problem 2 data

```{r reading_and_cleaning_p2_data, message = F}
# importing and cleaning the Mr. Trash Wheel excel sheet data
mr_tw_df =
  readxl::read_excel(
    paste0(
      "p8105_hw2_mts2203_files/",
      "202610 Trash Wheel Collection Data.xlsx"),
    sheet = "Mr. Trash Wheel",
    range = "A2:N743",
    na = c("", "NA", "."),
    skip = 1
  ) |>
  janitor::clean_names() |>
  filter(
    !is.na(dumpster)
  ) |>
  mutate(
    sports_balls = as.integer(round(sports_balls))
  )
```

## Reading, cleaning and organizing more Problem 2 data

```{r organizing_other_p2_data, message = F}
# importing and organizing the Professor Trash Wheel excel sheet data
prof_tw_df = 
  readxl::read_excel(
    paste0(
      "p8105_hw2_mts2203_files/",
      "202610 Trash Wheel Collection Data.xlsx"),
    sheet = "Professor Trash Wheel",
    range = "A2:M140",
    na = c("", "NA", "."),
    skip = 1
  ) |>
  janitor::clean_names() |>
  filter(
    !is.na(dumpster)
  )

# importing and organizing the Gwynnda Trash Wheel excel sheet data
gwynnda_tw_df = 
  readxl::read_excel(
    paste0(
      "p8105_hw2_mts2203_files/",
      "202610 Trash Wheel Collection Data.xlsx"),
    sheet = "Gwynnda the Good Wheel of the W",
    range = "A2:L395",
    na = c("", "NA", "."),
    skip = 1
  ) |>
  janitor::clean_names() |>
  filter(
    !is.na(dumpster)
  )
```

## Combining Problem 2 datasets into a single tidy dataset

```{r combining_all_tw_datasets, message = F}
# adding a variable identifying the Mr. Trash Wheel dataset
# standardizing the year and plastic_bottles variables
mr_tw_df = mr_tw_df |>
  mutate(
    year = as.numeric(year),
    plastic_bottles = as.numeric(plastic_bottles),
    trash_wheel = "Mr. Trash Wheel"
  )

# adding a variable identifying the Professor Trash Wheel dataset
# standardizing the year and plastic_bottles variables
prof_tw_df = prof_tw_df |>
  mutate(
    year = as.numeric(year),
    plastic_bottles = as.numeric(plastic_bottles),
    trash_wheel = "Professor Trash Wheel"
  )

# adding a variable identifying the Gwynnda Trash Wheel dataset
# standardizing the year and plastic_bottles variables
gwynnda_tw_df = gwynnda_tw_df |>
  mutate(
    year = as.numeric(year),
    plastic_bottles = as.numeric(plastic_bottles),
    trash_wheel = "Gwynnda the Good Wheel of the W"
  )

# combining all three datasets
tw_df = bind_rows(mr_tw_df, prof_tw_df, gwynnda_tw_df) |>
  relocate(trash_wheel)
```

## Paragraph about Problem 2 data

The combined `tw_df` dataset consists of `r nrow(tw_df)` observations. The key variables are `r paste(names(tw_df), collapse = ", ")`. Given the available data, the total weight of trash collected by Professor Trash Wheel is `r tw_df |> filter(trash_wheel == "Professor Trash Wheel") |> pull(weight_tons) |> sum(na.rm = TRUE)` tons. The total number of cigarette butts collected by Gwynnda in June of 2022 is  `r tw_df |> filter(trash_wheel == "Gwynnda the Good Wheel of the W", month == "June", year == 2022) |> pull(cigarette_butts) |> sum(na.rm = TRUE)`.


# Problem 3

## Reading and cleaning the Problem 3 data

``` {r reading_and_tidying_p3_data, message = F}  
# importing and tidying MCI_baseline CSV data
mci_baseline_df =
  read_csv(
    file = paste0(
      "p8105_hw2_mts2203_files/",
      "MCI_baseline.csv"),
    skip = 1,
    na = c("", "NA", ".")
  ) |>
  janitor::clean_names() |>
  mutate(
    sex = case_match(
      sex, 
      1 ~ "male",
      0 ~ "female"
    ),
    apoe4 = case_match(
      apoe4,
      1 ~ "carrier",
      0 ~ "non-carrier"
    )
  ) |>
  filter(
    age_at_onset > current_age | is.na(age_at_onset)
  )
```

## Discussing the MCI baseline demographics dataset import and features

The MCI_baseline.csv dataset was imported as `mci_baseline_df`. To import the CSV file, I used the `read_csv()` `tidy` function. To clean the data, I began by skipping the first line of the dataset. This line contained notes that were unnecessary to include considering the clarity of many column variable names. However, for those column variables without such clarity, I used the `mutate()` function, with `case_match()` nested within it, to recode their values; thus, amplifying their clarity. To ensure that all participants met the 'no MCI at baseline' criterion, I used the `filter()` function to isolate those whose `age_at_onset` exceeded `current_age`, as well as those who remained MCI free after follow-up, i.e., the NAs. In other efforts to clean this dataset, I standardized across variables several categories. These included the presentation of missing variables, and the presentation of column variables. The latter task was completed via the `clean_names()` function in the `janitor` package. The tidy `mci_amyloid_df`dataset has `r nrow(mci_baseline_df)` rows and `r ncol(mci_baseline_df)` columns. These are its column variables: `r paste(names(mci_baseline_df), collapse = ", ")`. In summary, `r nrow(mci_baseline_df)` participants were recruited, `r mci_baseline_df |> filter(!is.na(age_at_onset)) |> nrow()` of them developed MCI, `r mci_baseline_df |> summarise(mean_age = mean(current_age, na.rm = TRUE))` was the average baseline age, and `r mci_baseline_df |> filter(sex == "female") |> summarise(carrier = mean(apoe4 == "carrier", na.rm = TRUE))` was the proportion of women in the study who were APOE4 carriers.



## Reading and cleaning more Problem 3 data
```{r reading_and_tidying_biomarker_data, message = F}
# importing and tidying mci_amyloid CSV data
mci_amyloid_df =
  read_csv(
    file = paste0(
      "p8105_hw2_mts2203_files/",
      "mci_amyloid.csv"),
    skip = 1,
    na = c("", "NA", ".")
  ) |>
  janitor::clean_names() |>
  rename(
    id = study_id,
    baseline_years = baseline
  )
```

## Commenting on the mci_amyloid import process steps and dataset features

The mci_amyloid.csv dataset was imported as `mci_amyloid_df`. To import the CSV file, I used the `read_csv()` `tidy` function. To clean the data, I began by skipping the first line of the dataset. This line contained notes that were unnecessary to include considering the clarity of many column variable names. However, for those column variables without such clarity, I used the `rename()` function to change their values either for clarity of measurement or for uniformity with `mci_baseline_df`. In other efforts to clean this dataset, I standardized variables across several categories. These included the presentation of missing variables, and the presentation of column variables. The latter task was completed via the `clean_names()` function in the `janitor` package. The tidy `mci_amyloid_df`dataset has `r nrow(mci_amyloid_df)` rows and `r ncol(mci_amyloid_df)` columns. These are its column variables: `r paste(names(mci_amyloid_df), collapse = ", ")`.

## Checking participants between datasets

```{r checking_for_participants, message = F}
# checking for participants in mci_baseline but not mci_amyloid
mci_baseline_df |> 
  anti_join(mci_amyloid_df, by = "id") |> 
  nrow()

# checking for participants in mci_amyloid but not mci_baseline
mci_amyloid_df |> 
  anti_join(mci_baseline_df, by = "id") |> 
  nrow()
```

## Commenting on participant check finding

The `mci_baseline_df` dataset contains 8 participants who are not in the `mci_amyloid_df` dataset.
The `mci_amyloid_df` dataset contains 16 participants who are not in the `mci_baseline_df` dataset.

## Combining Problem 3 datasets

```{r combining_p3_datasets, message = F}
# combining mci_baseline_df and mci_amyloid_df to retain shared participants
mci_df =
  inner_join(mci_baseline_df, mci_amyloid_df, by = "id")
```

## Describing resulting dataset

The resulting `mci_df` dataset is a combination of the `mci_baseline_df` and the `mci_amyloid` datasets where only participants who appear in both datasets are retained. To do this, the `inner_join()` function was applied to both, with the `id` variable anchoring the datasets to one another. Altogether, the `mci_df` dataset has `r nrow(mci_df)` rows and `r ncol(mci_df)` columns. Its column variables include `r paste(names(mci_df), collapse = ", ")`.

## Exporting result as a CSV to my Github data directory

```{r exporting_combined_csv, message = F}
# exporting the resulting mci_df dataset as a CSV to my Github data directory
write_csv(mci_df, "p8105_hw2_mts2203_files/mci_df.csv")
```